---
title: "Recitation 1 – HWE"
author: "Rei Jia"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Step 1: Input Data
```{r}
# Genotype counts
AA <- 50
Aa <- 30
aa <- 20 #Question 1 - what is this value?

# Total population size
N <- AA + Aa + aa
N
```

Step 2: Calculate Observed Allele Frequencies
```{r}
# Observed allele frequencies

p <- (2 * AA + Aa) / (2 * N) # Frequency of A

q <- (2 * aa + Aa) / (2 * N) #Frequency of a # Question 2 - what is the equation?
```

Step 3: Expected Genotype Frequencies Under HWE
```{r}
# Expected genotype frequencies

exp_AA <- p^2

exp_Aa <- 2 * p * q

exp_aa <- q^2
```

Step 4: Expected Genotype Counts
```{r}
# Convert frequencies to counts

exp_counts <- c(AA = exp_AA * N, Aa = exp_Aa * N, aa = exp_aa * N)

##QUESTION 3: explain in words what is occurring on this line of code?
```

QUESTION 3: explain in words what is occurring on this line of code?

Step 5: Perform Chi-Square Test for HWE. In order to assess whether our population is statistically under Hardy-Weinberg Equilibrium, we will perform a statistical test called the “chi-squared.” This test compares expected and observed values and determines whether the values “fit together” (more explicitly whether they are drawn from the same or different probability distribution). A non-significant P-value indicates equilibrium, a significant one indicates a deviation from equilibrium.

Chi-squared test:
```{r}
obs_counts <- c(AA = AA, Aa = Aa, aa = aa)

chisq_test <- chisq.test(x = obs_counts, p = c(exp_AA, exp_Aa, exp_aa))

chisq_test

##QUESTION 4: Report and interpret the result of this analysis
```

Step 6: Visualize Observed vs. Expected Genotypes
```{r}
# Bar plot 
##QUESTION 5: put this graph in your report. Describe how the graph contextualizes the results of the test

barplot(rbind(obs_counts, exp_counts),

beside = TRUE, col = c("skyblue", "orange"),

names.arg = c("AA", "Aa", "aa"),

main = "Observed (blue) vs. Expected (orange)\nGenotype Counts",ylab = "Counts")
```
