---
title: "Recitation 6 – Integrating forces jointly with temporal fluctuating selection"
author: "Rei Jia"
date: "2026-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Step 1: Make a formula for seasonal selection (see Recitation 4)
```{r}
# Initial allele frequency
p <- 0.1

temporal_selection = function( p, sii, sij, sjj ){
  q=1-p
  num=(p^2)*(1-sii) + p*q*(1-sij) 
  dem=(p^2)*(1-sii) + 2*p*q*(1-sij) + (q^2)*(1-sjj)
  p_t1 = num/dem
  return(p_t1)
} 
```

Step 2: Make a formula for directional selection with dominance

<<QUESTION 1: Code a function for directional selection with dominance… you can code the formula for delta (or the rate) directly... Look at your notes>>

```{r}
directional_selection = function(p, s, h){
  q=1-p
  pq=p*q
  sii=0
  sij=1-h*s
  sjj=s
  #num=pq*(sjj*(p*h+q[1-h]))
  num=pq*(p*(h*sjj-sii) + q*(sjj-h*sjj))
  dem=1-((p^2)*sii) - (2*p*q*sij) - ((q^2)*sjj)
  delta_p = num / dem
  return(delta_p)

}
```

Step 3: Make a formula for mutation
<<QUESTION 2: Code a function for mutational effects (ignore back mutation)… look at your lecture notes>>

```{r}
mutation = function( p, u ){
  q=1-p
  p_t1 = p*(1-u)
  return(p_t1)
}
```

Step 4: Create a simulation framework for these forces

```{r}
p_recur=c()

#lets simulate 100 generations
for(g in 1:100){
  
  if(g==1){
    p_recur[g]=p
  } else{
    #seasonal selection
    if(grepl("[0-4]$", g)){
      print("summer")
      sii=0.0; sij=0.2; sjj=0.4 
    }else if(grepl("[5-9]$", g)){
      print("winter")
      sii=0.4; sij=0.2; sjj=0.0 
    }
    seasonal_component = temporal_selection(p_recur[g-1], sii, sij, sjj)
    seasonal_delta = seasonal_component - p_recur[g-1]
    
    #directional component → you should have created a formula that estimates the delta directly!
    s=0.1; h=0.3 
    directional_delta = directional_rate(p_recur[g-1], s, h)
    
    
    #mutational component ### ignore back mutation for now!
    u=1e-8
    mutational_component = mutation(p_recur[g-1], u)
    mutational_delta = mutational_component - p_recur[g-1]
    
    ### Evaluate the evolutionary change
    
    p_recur[g] = p_recur[g-1] + 
      seasonal_delta +  
      directional_delta + 
      mutational_delta
    
  } # close if-else
} # close loop

#create a dataframe to save your results
integrated_selection = data.frame(g=1:100, p_recur)
```

Step 5: Plot the simulations
```{r}
plot(integrated_selection$g, integrated_selection$p_recur, type ='l',ylim=c(0,1), ylab="Allele Freq", xlab="Generation")
```

<<QUESTION 3: Print this graph in your report and describe what is happening. What is the long term evolutionary fate of these alleles? What inferences can you make about the impact of “multi-stressors” in adaptive evolution in wild populations>>

Step 6: a small challenge
Mutation rate doesn't seem to be impacting the simulation much… lets see what would happen if we were to introduce a rate of back-mutation into the system (v)... Modify your function for mutation rate to include back mutation:

```{r}
mutation = function( p, u, v ){
q=1-p
p_t1 = p*(1-u)+ (1-p)*v
return(p_t1)
}
```

<<QUESTION 4: holding v constant (v=1x10^-8) what magnitude of mutation rates (μ) would you need to have to notice a difference… A ballpark is fine, but try to “force” the simulation to a steady state of p=0.9... provide a plot to showcase your point (you may want to simulate 1000 generations…) >>.

```{r}
temporal_selection = function( p, sii, sij, sjj ){
  q=1-p
  num=(p^2)*(1-sii) + p*q*(1-sij) 
  dem=(p^2)*(1-sii) + 2*p*q*(1-sij) + (q^2)*(1-sjj)
  p_t1 = num/dem
  return(p_t1)
} 

directional_selection = function(p, s, h){
  q=1-p
  pq=p*q
  sii=0
  sij=1-h*s
  sjj=s
  num=pq*(p*(h*sjj-sii) + q*(sjj-h*sjj))
  dem=1-((p^2)*sii) - (2*p*q*sij) - ((q^2)*sjj)
  delta_p = num / dem
  return(delta_p)

}

mutation = function( p, u, v ){
q=1-p
p_t1 = p*(1-u)+ (1-p)*v
return(p_t1)
}

simulate = function(u, v, gens=1000){
  p_recur = numeric(gens)
  p_recur[1] = 0.1
  
  for(g in 2:gens){
    
    # Seasonal selection
    if(g %% 10 < 5){
      sii=1.0; sij=0.8; sjj=0.6  # summer
    } else{
      sii=0.6; sij=0.8; sjj=0.1  # winter
    }
    
    seasonal = temporal_selection(p_recur[g-1], sii, sij, sjj)
    seasonal_delta = seasonal - p_recur[g-1]
    
    # Directional selection
    directional_delta = directional_rate(p_recur[g-1], 0.1, 0.3)
    
    # Mutation
    mut = mutation(p_recur[g-1], u, v)
    mut_delta = mut - p_recur[g-1]
    
    # Update
    p_recur[g] = p_recur[g-1] + seasonal_delta + directional_delta + mut_delta
  }
  
  return(p_recur)
}

# Run simulations
u_values = c(1e-8, 1e-4, 1e-3, 1e-2)
v = 1e-8

results = lapply(u_values, function(u) simulate(u, v))

# Plot
plot(results[[1]], type='l', col='black', ylim=c(0,1),
     ylab="Allele frequency (p)", xlab="Generation",
     main="Mutation vs Selection Balance")

lines(results[[2]], col='blue')
lines(results[[3]], col='green')
lines(results[[4]], col='red')

abline(h=0.9, lty=2)

legend("bottomright",
       legend=c("u=1e-8","u=1e-4","u=1e-3","u=1e-2"),
       col=c("black","blue","green","red"),
       lty=1)
```